\documentclass[9pt]{article}

\usepackage{xargs}
\usepackage{mathpartir}
\usepackage{xifthen}
\usepackage{upgreek}
\usepackage{amsfonts}

\input{defs}
\input{figs}

\begin{document}

  \SyntaxFigure

  \SemanticsFigure

\clearpage 

  The NewChan rule simply adds a new channel to the global channel
  store, and returns a new identifier for it. Note that the semantics of
  the language currently lack a binding form that actually allows this
  identifier to be used (since threads require the identifier), but such
  an extension is trivial. 

  The Comm rule allows two processes to communicate synchronously, if
  the times on the channel are all in sync. This indicates that none of
  the threads involved are currently trying to backtrack. It updates the
  channels times to be the max of the sender or receivers time plus 1,
  saves the current continuation to the continuation stack, and updates
  the time of both threads. Additionally, it adds the channels and their
  time to the lists of communicated channels, which are needed to ensure
  partners backtrack with the thread.

  Choose creates a new continuation whose stable time is the stable time
  prior to the new choice point, and whose continuation is the second
  choice. This ensure that backtracking once takes the second choice,
  and backtracking again goes to the prior stable time.

  Yield simply forces a context switch if the system cannot support all
  threads running in parallel. Of course, a scheduler would also be
  required in such a situation, which is not formalized here.

  Backtrack finds the continuation whose logical time matches the stable
  time of the thread, restores the timestamps on all the
  channels it has used to the times stored for each in the past
  continuation.  Finally, it transfers to a 'sync' context, which waits
  for neighbors to backtrack and resynchronized the channel times.  
  Note: if new channels have been created between the current context
  and the last stable time, this doesn't currently handle them. I'm not
  entirely sure how to handle such channels just yet, but this situation
  can be avoided and the system still be useful.

  Return-End provides the conditions for terminating. Both the sender
  and receiver on all channels the thread is involved in must 'release'
  the channel. This occurs by each setting their time on the channel to
  some maximal time value, $\maxtime$

  Return-Wait updates all the channels the thread has used with the
  $\maxtime$ value, and continues to wait for something in the global
  context to change before it can make progress.

  Return-Back provides for backtracking of neighbor threads. The thread
  must wait on all channels it has used, and if the threads partner has set
  it's time backwards (that is, the neighbors time is less than the
  channels time), then the thread must backtrack and join it's partner.

  Finally, Sync, as mentioned previously, allows a backtracking thread
  to wait for it's neighbors to join it at a previous logical time,
  before resynchronizing the channel.


  The formal language lacks $par$, functions, binding forms, and some
  other uninteresting details. However, what's here should provide
  enough interesting semantics for now.
\end{document}
